Survival Compass: Statistical Insights into Lung Cancer Patients' Journey Post-Diagnosis

### 2024-04-05 paper/paper.qmd
updating paper - working on modeling

# load packages
library(ggplot2)
library(dplyr)
library(showtext)

# read in data
analysis_data <- read.csv(file = here::here("data/analysis_data/lung_cancer_analysis_data.csv"))

@@

# create bar chart for pathogenic_stage
pathogenic_stage_counts <- analysis_data %>%
  group_by(pathogenic_stage) %>%
  summarise(count = n ()) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  arrange(desc(pathogenic_stage))

ggplot(pathogenic_stage_counts, aes(x = pathogenic_stage, y = percentage)) +
  geom_bar(stat = "identity", fill = "lightpink") +
  labs(title = "Percentage of Patients by Pathogenic Stage",
       x = "Pathogenic Stage",
       y = "Percentage") +
  theme_minimal() +
  theme(
    text = element_text(family = "patrick") # Set font to Patrick Hand
  )

@@

# create bar chart for treatment decision
treatment_decision_counts <- analysis_data %>%
  group_by(treatment_decision) %>%
  summarise(count = n ()) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  arrange(desc(treatment_decision))

ggplot(treatment_decision_counts, aes(x = reorder(treatment_decision, -percentage), y = percentage)) +
  geom_bar(stat = "identity", fill = "blue") +
  geom_text(aes(label = sprintf("%.1f%%", percentage)), hjust = -0.2, color = "black", size = 3) +
  labs(title = "Percentage of Patients by Treatment Decision",
       x = "Treatment Decision",
       y = "Percentage") +
  theme_minimal() +
  coord_flip() +
  theme(
    text = element_text(family = "patrick") # Set font to Patrick Hand
  )

@@

Talk way more about it.

@@

The goal of our modeling strategy is twofold. Firstly, we aim to investigate the post-diagnosis survival outcomes of lung cancer patients, considering the pathogenic stage. Secondly, we seek to assess the efficacy of two common treatment modalities, radiation therapy and chemotherapy, in improving survival chances across different pathogenic stages.

Here, we briefly describe the Bayesian analysis model used to investigate survival outcomes in lung cancer patients based on pathogenic stage and treatment modality. Detailed background information and diagnostic assessments are provided in the [Appendix -@].

Our Bayesian analysis model is tailored to capture the complex relationships between pathogenic stage, treatment modality, and post-diagnosis survival outcomes in lung cancer patients. The model accounts for the multifaceted nature of cancer progression, incorporating tumor size, metastatic spread, and lymph node involvement to characterize pathogenic stage.

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

@@

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

The Bayesian analysis model serves as a powerful tool for investigating post-diagnosis survival outcomes in lung cancer patients. By incorporating pathogenic stage information and treatment modalities into the analysis, we aim to elucidate the factors influencing patient prognosis and provide insights to guide personalized treatment decisions.

@@

# Load required libraries
library(ggplot2)

# Scatter plot survival time vs. pathogenic stage - to compare survival probabilities
ggplot(analysis_data, aes(x = pathogenic_stage, y = days_to_death)) +
  geom_point(size = 3, color = "lightpink") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  
  # Add a linear trend line
  labs(title = "Survival Time vs. Pathogenic Stage Post-Diagnosis",
       x = "Pathogenic Stage",
       y = "Survival Time (days)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  # Rotate x-axis labels for better readability
  theme(text = element_text(family = "patrick") 
  # Set font to Patrick Hand
  )

# Line graph with trend lines 
ggplot(analysis_data, aes(x = pathogenic_stage, y = days_to_death)) +
  geom_line(color = "blue", group = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  
  # Add a linear trend line
  labs(title = "Days to Death vs. Pathogenic Stage for Lung Cancer Patients",
       x = "Pathogenic Stage",
       y = "Days to Death") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

# not sure what this is 
# Plotting
ggplot(analysis_data, aes(x = days_to_death, fill = pathogenic_stage)) +
  geom_density(alpha = 0.5) +
  geom_density(color = "black", size = 1.5) +
  labs(title = "Density Plot of Days to Death by Pathogenic Stage for Lung Cancer Patients",
       x = "Days to Death",
       y = "Density") +
  scale_fill_manual(values = c("lightgreen", "lightgreen", "lightgreen","forestgreen", "forestgreen", "forestgreen", "darkgreen","darkgreen", "darkgreen", "red")) +
  theme_minimal()

# not sure .2
ggplot(analysis_data, aes(x = days_to_death, fill = pathogenic_stage)) +
  geom_density(alpha = 0.5) +
  geom_density(color = "black", size = 1.5) +
  labs(title = "Density Plot of Days to Death by Pathogenic Stage for Lung Cancer Patients",
       x = "Days to Death",
       y = "Density") +
  scale_fill_manual(values = c("lightgreen", "lightgreen", "lightgreen","forestgreen", "forestgreen", "forestgreen", "darkgreen","darkgreen", "darkgreen", "red")) +
  theme_minimal() +
  facet_wrap(~pathogenic_stage, scales = "free")

# Boxplot 
ggplot(analysis_data, aes(x = pathogenic_stage, y = days_to_death)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Survial Distribution Post-Diagnosis by Pathogenic Stage",
       x = "Pathogenic Stage",
       y = "Days to Death") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

@@

# Scatter plot survival time vs. Treatment Decision
ggplot(analysis_data, aes(x = treatment_decision, y = days_to_death)) +
  geom_point(size = 3, color = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  
  # Add a linear trend line
  labs(title = "Survival Time vs. Treatment Decision Post-Diagnosis",
       x = "Treatment Decision",
       y = "Survival Time (days)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  # Rotate x-axis labels for better readability
  theme(text = element_text(family = "patrick") 
  # Set font to Patrick Hand
  )
#########################

# create bar chart for treatment decision
treatment_decision_counts <- analysis_data %>%
  group_by(treatment_decision) %>%
  summarise(count = n ()) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  arrange(desc(treatment_decision))

ggplot(treatment_decision_counts, aes(x = reorder(treatment_decision, -percentage), y = percentage)) +
  geom_bar(stat = "identity", fill = "blue") +
  geom_text(aes(label = sprintf("%.1f%%", percentage)), hjust = -0.2, color = "black", size = 3) +
  labs(title = "Percentage of Patients by Treatment Decision",
       x = "Treatment Decision",
       y = "Percentage") +
  theme_minimal() +
  coord_flip() +
  theme(
    text = element_text(family = "patrick") # Set font to Patrick Hand
  )

@@

#######################

>>>>

# load packages
library(ggplot2)
library(dplyr)
library(showtext)
library(arrow)

### read in data ###
analysis_data <- read_parquet(file = here::here("data/analysis_data/lung_cancer_analysis_data.parquet"))

@@

# create bar chart for pathologic_stage
pathologic_stage_counts <- analysis_data %>%
  group_by(pathologic_stage) %>%
  summarise(count = n ()) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  arrange(desc(pathologic_stage))

ggplot(pathologic_stage_counts, aes(x = pathologic_stage, y = percentage)) +
  geom_bar(stat = "identity", fill = "lightpink") +
  labs(title = "Percentage of Patients by Pathologic Stage",
       x = "Pathologic Stage",
       y = "Percentage") +
  theme_minimal() +
  theme(
    text = element_text(family = "patrick") # Set font to Patrick Hand
  )

@@

@@

@@

## Linear Regression Model 
# interested in predicting days_to_death, a continuous outcome variable based on
# multiple predictor variables  (pathogenic_stage, lymph_node_involvement, 
# presence_of_distant_metastasis, tumor_size, treatment_decision, treatment_type)
# Linear regression model are suitable for predicting continuous outcomes
# quantifying relationship between predictor variables and outcome variable.

# Fit a linear regression model to predict continuous outcome variable 
# 'days_to_death'

## Model set-up
In this study, we aim to investigate the survival outcomes of lung cancer patients post-diagnosis. Let $y_i$ denote the number of days from diagnosis to death for patient $i$. We consider several predictors including pathogenic stage, lymph node involvement, presence of distant metastasis, tumor size, and treatment type.

We model the survival time $y_i$ using a linear regression framework where $mi_i$ is the expected survival time for patient $i$, and $σ$ represents the standard deviation of the survival times. The linear predictor $mu_i$ is specified as:

\begin{align*}
y_i | \mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_{\text{pathologic\_stage}} \times \text{pathologic\_stage}_i + \beta_{\text{lymph\_node}} \times \text{lymph\_node\_involvement}_i \\
&\quad + \beta_{\text{metastasis}} \times \text{presence\_of\_distant\_metastasis}_i + \beta_{\text{tumor\_size}} \times \text{tumor\_size}_i \\
&\quad + \beta_{\text{treatment\_type}} \times \text{treatment\_type}_i
\end{align*}

Here, \alpha represents the intercept term, \beta coefficients denote the effect of respective predictors, and subscripts $i$ denote individual patients.

Priors are specified for the model parameters as follows:

\begin{align*}
\alpha &\sim \text{Normal}(0, 2.5) \\
\beta_{\text{pathologic_stage}}, \beta_{\text{lymph_node}}, \beta_{\text{metastasis}}, \beta_{\text{tumor_size}}, \beta_{\text{treatment_type}} &\sim \text{Normal}(0, 2.5) \\
\sigma &\sim \text{Exponential}(1)
\end{align*}

TEMPLATE:
Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align}

\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

@@

We anticipate that the survival time of lung cancer patients post-diagnosis will be influenced by various clinical factors such as pathologic stage, extent of lymph node involvement, presence of distant metastasis, tumor size, and treatment type. Specifically, we expect that advanced pathologic stages, increased lymph node involvement, presence of distant metastasis, larger tumor sizes, and certain treatment types will be associated with shorter survival times.

The linear regression model allows us to quantify the relationships between these predictors and survival outcomes, providing valuable insights into the factors influencing the prognosis of lung cancer patients.

@@

# Load required libraries
library(ggplot2)
library(arrow)

### read in data ###
analysis_data <- read_parquet(file = here::here("data/analysis_data/lung_cancer_analysis_data.parquet"))

# Scatter plot survival time vs. pathologic stage - to compare survival probabilities
ggplot(analysis_data, aes(x = pathologic_stage, y = days_to_death)) +
  geom_point(size = 3, color = "lightpink") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  
  # Add a linear trend line
  labs(title = "Survival Time vs. Pathologic Stage Post-Diagnosis",
       x = "Pathologic Stage",
       y = "Survival Time (days)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  # Rotate x-axis labels for better readability
  theme(text = element_text(family = "patrick") 
  # Set font to Patrick Hand
  )

# Line graph with trend lines 
ggplot(analysis_data, aes(x = pathologic_stage, y = days_to_death)) +
  geom_line(color = "blue", group = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  
  # Add a linear trend line
  labs(title = "Days to Death vs. Pathologic Stage for Lung Cancer Patients",
       x = "Pathologic Stage",
       y = "Days to Death") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

# not sure what this is 
# Plotting
ggplot(analysis_data, aes(x = days_to_death, fill = pathologic_stage)) +
  geom_density(alpha = 0.5) +
  geom_density(color = "black", size = 1.5) +
  labs(title = "Density Plot of Days to Death by Pathologic Stage for Lung Cancer Patients",
       x = "Days to Death",
       y = "Density") +
  scale_fill_manual(values = c("lightgreen", "lightgreen", "lightgreen","forestgreen", "forestgreen", "forestgreen", "darkgreen","darkgreen", "darkgreen", "red")) +
  theme_minimal()

# not sure .2
ggplot(analysis_data, aes(x = days_to_death, fill = pathologic_stage)) +
  geom_density(alpha = 0.5) +
  geom_density(color = "black", size = 1.5) +
  labs(title = "Density Plot of Days to Death by Pathologic Stage for Lung Cancer Patients",
       x = "Days to Death",
       y = "Density") +
  scale_fill_manual(values = c("lightgreen", "lightgreen", "lightgreen","forestgreen", "forestgreen", "forestgreen", "darkgreen","darkgreen", "darkgreen", "red")) +
  theme_minimal() +
  facet_wrap(~pathologic_stage, scales = "free")

# Boxplot 
ggplot(analysis_data, aes(x = pathologic_stage, y = days_to_death)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Survial Distribution Post-Diagnosis by Pathologic Stage",
       x = "Pathologic Stage",
       y = "Days to Death") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

@@

#########################

@@
